{memer}: an R interface to the MEME Suite

### 2020-07-23 manuscript.Rmd
added abstract for bioc2020

>>>>

# Bioconductor Abstract (300 words)

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools implementing these features exist within
Bioconductor, they often rely on proprietary data types with poor
interoperability with other tools. Furthermore, several of these packages will
be deprecated in the next Bioconductor release, leaving a gap in functionality
with no clear alternatives.

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memer`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memer` uses standard R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and `GRanges`. `memer` outputs also
function as inputs to other `memer` commands, allowing simple construction of motif
analysis pipelines.

In addition to wrapping MEME Suite utilities, `memer` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. `memer` also provides
simple utilities for data visualization.

`memer` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memer`
allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

# Abstract Draft
Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing,comparison, and matching. Here we present `memer`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memer` uses base R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memer` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines.

In addition to wrapping MEME Suite utilities, `memer` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue
of interest. `memer` also provides simple utilities for data visualization where
appropriate.

`memer` is designed for maximum flexibility and ease of use to allow users to
iterate rapidly during analysis. Here we present several examples of how `memer`
allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R package landscape.
